Poincar\'e--Friedrichs constants over triangulated domains

### 2024-06-09 main.tex
Update on Overleaf.

We gather notions of simplicial meshes for the subsequent discussion. 
Apart from basic definitions and notational conventions, we pay special attention to the question of when a simplicial complex is a triangulation of a manifold. 
A particularly important class of simplicial complexes are so-called \emph{shellable} complexes. Such simplicial complexes are constructed by successively adding simplices in a particularly well-behaved manner. While shellable complexes are a very special class of complexes, they facilitate achieving the main goal of our exposition.
Our main references on simplicial complexes and shellability are the monographs by Ziegler~\cite{ziegler2012lectures} and Kozlov~\cite{kozlov2008combinatorial}.

\subsection{Basic notions}

A ${k}$-dimensional \emph{simplex} $T$ is the convex hull of ${k}+1$ affinely independent points $v_0, v_1, \ldots, v_{{k}} \in \mathbb{R}^{n}$. We call these the \emph{vertices} of the simplex $T$. 
If $S$ is a simplex whose vertices are also vertices of another simplex $T$, in which case $S \subseteq T$, 
then we call $S$ a subsimplex of $T$ and we call $T$ a supersimplex of $S$.

A family of simplices $\calT$ is a \emph{simplicial complex} or \emph{triangulation}, if its member simplices satisfy the following condition: 
(i) every subsimplex of any simplex in $\calT$ is itself part of $\calT$, and (ii) the intersection between any two simplices $T, T' \in \calT$ is either empty or a shared subsimplex of both $T$ and $T'$. 
The set of $k$-dimensional simplices in $\calT$ is denoted as $\subsimplex_{{k}}(\calT)$. 
We say that $\calT$ has dimension $n$ if it contains an $n$-dimensional simplex as a supersimplex for each simplex within it.

Given any simplex $T$, we write $\subsimplex(T)$ for the simplicial complex that contains all subsimplices of $T$. 
More specifically, $\subsimplex_{{k}}(T) \subseteq \subsimplex(T)$ denotes the set of $k$-dimensional subsimplices of $T$. 
Whenever $\calT$ is a simplicial complex and $T \in \calT$, we let $\supsimplex(\calT,T)$ be the set of simplices in $\calT$ that contain $T$.
We write $\supsimplex_{k}(\calT,T)$ for the set of $k$-dimensional simplices in $\supsimplex(\calT,T)$. 
Additionally, the notations $\Vertices(\calT) := \subsimplex_{0}(\calT)$ and $\Edges(\calT) := \subsimplex_{1}(\calT)$ refer to the vertices and edges of the simplicial complex, respectively. Similarly, if $\calT$ is an $n$-dimensional simplicial complex, we write $\Faces(\calT) := \subsimplex_{n-1}(\calT)$ for the faces of the triangulation.

When $\calT$ is a triangulation and $T \in \calT$, then $\patch_{\calT}(T)$ denotes the \emph{local patch} of $T$, which is the smallest simplicial subcomplex of $\calT$ that contains all supersimplices of $T$. Formally,
\begin{gather*}
    \patch_{\calT}(T) := \bigcup_{ T' \in \supsimplex(\calT,T) } \subsimplex(T').
\end{gather*}
A crucial observation is the following.

\begin{lemma}
 Let $\calT$ be an $n$-dimensional simplicial complex and let $T,T' \in \calT$.
 Then either $\calT_{n}(T) \cap \calT_{n}(T') = \emptyset$ or there exists $S \in \calT$
 such that $\calT_{n}(T) \cap \calT_{n}(T') = \calT_{n}(S)$ with $T, T' \subseteq S$.
\end{lemma}
\begin{proof}
 Let $F \in \calT$ be $n$-dimensional.
 We have $F \in \patch_{\calT}(T )$ if and only if all vertices of $T $ are vertices of $F$.
 We have $F \in \patch_{\calT}(T')$ if and only if all vertices of $T'$ are vertices of $F$.
 Consequently, $F \in \patch_{\calT}(T) \cap \patch_{\calT}(T)$ if and only if $F \in \patch_{\calT}(S)$,
 where $S \in \calT$ and $S$ is the convex closure of $F$ and $F'$.
\end{proof}

Given an $n$-dimensional simplicial complex $\calT$, 
we call $n$-simplices $S,T \in \calT$ \emph{strongly connected in $\calT$} if there exists a sequence $S_0=S,S_1,\dots,T=S_m \in \calT$ such that $S_{i} \cap S_{i-1}$ is a face of both $S_{i}$ and $S_{i-1}$ for all $1 \leq i \leq m$. Clearly, strongly connected in $\calT$ is an equivalence relation. A \emph{strongly connected component} of $\calT$ is an equivalence class under this equivalence relation, and we call $\calT$ strongly connected if it has only one strongly connected component.

@@

\section{Shellable triangulated domains}

% 2. Develop manifold-like complexes and characterize them 
We call an $n$-dimensional simplicial complex $\calT$ \emph{non-branching} if every $F \in \calT$ of dimension $n-1$ is contained in at most two $n$-simplices of $\calT$. We call those $F \in \calT$ of dimension $n-1$ contained within only one parent a \emph{boundary face}.
We write $\partial\calT$ for the simplicial subcomplex of $\calT$ of dimension $n-1$ 
whose full-dimensional members are exactly the boundary faces of $\calT$.

An $n$-dimensional simplicial complex is called \emph{manifold-like} if for every $x \in \bigcup \calT$
we have one of the following two conditions:
\begin{itemize}
 \item 
 there exists $\epsilon > 0$ 
 and a homeomorphism 
 \begin{align*}
    \phi : \Ball_\epsilon(x) \cap \bigcup\calT  
    \rightarrow 
    \left\{ x \in \bbR^{n} \suchthat |x| \leq 1 \right\}
 \end{align*}
 such that $\phi(x) = 0$,
 \item 
 there exists $\epsilon > 0$ and a homeomorphism 
 \begin{align*}
    \phi : \Ball_\epsilon(x) \cap \bigcup\calT 
    \rightarrow 
    \left\{ x \in \bbR^{n} \suchthat |x| \leq 1, x_n \leq 0 \right\}
 \end{align*}
 such that $\phi(x) = 0$.
\end{itemize}
This characterizes the union of the simplicial complex $\calT$ as a topological manifold with boundary.

We would like to have a purely combinatorial and computable criterion that characterizes manifold-like simplicial complexes. A first step in that direction is the following result.

\begin{lemma}\label{lemma:characterizationofmanifoldcomplexes}
    Let $\calT$ be an $n$-dimensional simplicial complex. The following are equivalent:
    \begin{itemize}
     \item $\calT$ is manifold-like
     \item $\patch_{\calT}(V)$ is the triangulation of a ball for all $V \in \Vertices(\calT)$
     \item $\patch_{\calT}(V)$ is manifold-like for all $V \in \Vertices(\calT)$
     \item $\patch_{\calT}(S)$ is the triangulation of a ball for all $S \in \calT$
     \item $\patch_{\calT}(S)$ is manifold-like for all $S \in \calT$
    \end{itemize}
\end{lemma}
\begin{proof}
    \color{red} TODO
\end{proof}

\begin{remark}
    Clearly, not every simplicial complex is the triangulation of some (embedded) topological manifold with or without boundary. 
    Conversely, for each dimension $n \geq 5$ there exists an $n$-dimensional compact topological manifold without a triangulation. % TODO: are these compact?
    % TODO There are computational limits to algorithmically deciding whether a simplicial complex represents a manifold. 
\end{remark}

@@

    \caption{Example of a connected $2$-dimensional manifold-like simplicial complex, which triangulates an annulus, that does neither admit a manifold-like $0$-patching nor manifold-like $1$-patching.}\label{figure:annuluscounterexample}
\end{figure}

@@

We turn our attention to a different idea of how to incrementally construct a simplicial complex while maintaining well-behaved intermediate complexes. 
Given an $n$-dimensional simplicial complex $\calT$, a \emph{shelling} is an enumeration of the $n$-simplices $T_{0}, T_{1}, T_{2}, \dots \in \subsimplex_{n}(\calT)$ such that $( \patch_{\calT}(T_{0}) \cup \patch_{\calT}(T_{1}) \cup \dots \cup \patch_{\calT}(T_{m}) ) \cap \patch_{\calT}(T_{m+1})$ is a simplicial complex of dimension $n-1$. Notice that the last intersection is automatically a subcomplex of the boundary complex of the last simplex $T_{m}$, and is thus manifold like.

% \begin{lemma}
%     If an $n$-dimensional simplicial complex $\calT$ has a shelling $T_{0}, T_{1}, T_{2}, \dots$, then it is manifold-like.
%     In particular, $\patch_{\calT}(T_{0}) \cup \patch_{\calT}(T_{1}) \cup \dots \cup \patch_{\calT}(T_{m})$ is manifold-like for every $m \geq 0$.
% \end{lemma}
% \begin{proof}    
% \end{proof}

We notion of shelling is taken from Ziegler's book~\cite{ziegler2012lectures}

@@

%

A simplicial $n$-dimensional complex $\calT$ is called shellable if its $n$-simplices can be arranged in linear order $T_0, T_1, \dots$ in such a way that the subcomplex $( \cup_{i=0}^{k-1} T_i ) \cap T_k$ has dimension $n-1$ for all $k > 0$.

\begin{remark}
    A shellable simplicial complex that triangulates a manifold has a very restricted topology. 
    The Mayer-Vietoris theorem implies restrictions on the homology groups of simplicial complexes iteratively constructed via shelling.

@@

    % then $H_{n}(\calT') \simeq H_{n}(\calT \oplus T)$ and $H_{n-1}(\calT') \simeq H_{n-1}(\calT \oplus T)$ follow immediately.
    % Moreover, Thus $H_{n}(\calT') = H_{n}(\calT) \oplus H_{n}(\calS)$.
    % We conclude that any shellable simplicial complex has the homology groups of a disk or a sphere. 
    % NB: the zeroth homology is always free, and the n-th homology group, being the kernel of a mapping free abelian group, is free as well.
    % \begin{align*}
    %     \begin{CD}
    %         0 \longrightarrow H_{1}(\calT) \oplus H_{1}(T) \longrightarrow H_{1}(\calT') \longrightarrow H_{0}(\calS) \longrightarrow H_{0}(\calT) \oplus H_{0}(T) \longrightarrow H_{0}(\calT') \longrightarrow 0 
    %     \end{CD}
    % \end{align*}
\end{remark}

>>>>

We gather basic notions and definitions concerning simplicial meshes.

A ${k}$-dimensional \emph{simplex} $T$ is the convex hull of ${k}+1$ affinely independent points $v_0, v_1, \ldots, v_{{k}} \in \mathbb{R}^{n}$. We call these the \emph{vertices} of the simplex $T$. 
If $S$ is a simplex whose vertices are also vertices of another simplex $T$, in which case $S \subseteq T$, 
then we call $S$ a \textit{subsimplex} of $T$ and we call $T$ a \textit{supersimplex} of $S$.

The strictly positive convex combinations of the simplex' vertices are the \textit{interior} of the simplex,
and its remaining points constitute the \textit{boundary} of the simplex.
\\

A finite family of simplices $\calT$ is a \emph{simplicial complex} or \emph{triangulation}, if it satisfies the following conditions: 
(i) $\calT$ contains all the subsimplices of its members (ii) any non-empty intersection of two members of $\calT$ is a common subsimplex of each of them. 
% (i) every subsimplex of any simplex in $\calT$ is itself part of $\calT$, and (ii) the intersection between any two simplices $T, T' \in \calT$ is either empty or a shared subsimplex of both $T$ and $T'$. 
The set of $k$-dimensional simplices in $\calT$ is denoted as $\subsimplex_{{k}}(\calT)$. 
We say that a simplicial complex $\calT$ has \textit{dimension $n$} or is \textit{$n$-dimensional} if each of its simplices is a subset of an $n$-dimensional member of that complex.\footnote{We warn the reader that simplicial complexes that we call $n$-dimensional are called purely $n$-dimensional in the literature on polytopes.} 
We also write $\underlying{\calT}$ for the \textit{underlying set} of the simplicial complex $\calT$, which is defined as the union 
\[
    \underlying{\calT} := \bigcup_{ T \in \calT } T
    .
\]
We call a set \textit{triangulable} if it is the underlying set of some triangulation.

\begin{remark}
    Our main interest in this manuscript are finite triangulations of compact sets. 
    There is another reason why we insist on the triangulation being finite: 
    if we do not require our simplicial complexes to be finite,
    then our definition would be considerably more permissive than the typical definitions in topology. 
    % 
    For example, if we allowed simplicial complex to be infinite, 
    then any subset of Euclidean space would give rise to an ``infinite $0$-dimensional complex''. But this obviously does not reflect the topology. 
    %
    A more interesting such example is this:
    the Cantor set $\calC \subseteq [0,1]$ is a compact set 
    that underlies an ``infinite $0$-dimensional simplicial complex''.
    If we place $\calC$ on the x-axis of a 2D coordinate system 
    and connect each member of $\calC$ to $(0,1)$ via a straight line segment, 
    then the resulting set is still compact, even path-connected,  
    but has an ``infinite $1$-dimensional triangulation''.
    % 
    However, none of these infinite families are simplicial complexes in the sense of topology~\cite{spanier1966homology}. 
\end{remark}

Given any simplex $T$, we write $\subsimplex(T)$ for the simplicial complex that contains all subsimplices of $T$. 
More specifically, $\subsimplex_{{k}}(T) \subseteq \subsimplex(T)$ denotes the set of $k$-dimensional subsimplices of $T$. 
Whenever $\calT$ is a simplicial complex and $T \in \calT$, we let $\supsimplex(\calT,T)$ be the set of simplices in $\calT$ that contain $T$.
We write $\supsimplex_{k}(\calT,T)$ for the set of $k$-dimensional simplices in $\supsimplex(\calT,T)$. 
Additionally, the notations $\Vertices(\calT) := \subsimplex_{0}(\calT)$ and $\Edges(\calT) := \subsimplex_{1}(\calT)$ refer to the vertices and edges of the simplicial complex, respectively. Similarly, if $\calT$ is an $n$-dimensional simplicial complex, we write $\Faces(\calT) := \subsimplex_{n-1}(\calT)$ for the faces (codimension $1$ members) of this triangulation.

\begin{remark}
    Our use of the term \textit{face} as is common in the finite element literature
    and is synonymous with \textit{facet} as used in the literature on polyhedral combinatorics.
    Notably, this terminology difers from the uses \textit{face} and \textit{facet} in the theory of polyhedra. 
\end{remark}

When $\calT$ is a triangulation and $T \in \calT$, then $\patch_{\calT}(T)$ denotes the \emph{local patch} or \textit{star} of $T$, 
which the simplicial subcomplex of $\calT$ that contains all supersimplices of $T$ and their subsimplices. Formally,
\begin{gather*}
    \patch_{\calT}(T) := \bigcup_{ T' \in \supsimplex(\calT,T) } \subsimplex(T').
\end{gather*}
A crucial observation is the following.

\begin{lemma}
 Let $\calT$ be an $n$-dimensional simplicial complex and let $S,S' \in \calT$.
 Then either $\calT_{n}(S) \cap \calT_{n}(S') = \emptyset$ or there exists $S'' \in \calT$
 such that $\calT_{n}(S) \cap \calT_{n}(S') = \calT_{n}(S'')$ with $S, S' \subseteq S''$.
\end{lemma}
\begin{proof}
 Let $T \in \calT$ be $n$-dimensional.
 We have $T \in \patch_{\calT}(S )$ if and only if all vertices of $S $ are vertices of $T$.
 We have $T \in \patch_{\calT}(S')$ if and only if all vertices of $S'$ are vertices of $T$.
 Consequently, $T \in \patch_{\calT}(S) \cap \patch_{\calT}(S')$ if and only if $T \in \patch_{\calT}(S'')$,
 where $S'' \in \calT$ is the convex closure of $S$ and $S'$.
\end{proof}

We will use a specific notion of connectivity. 
Given an $n$-dimensional simplicial complex $\calT$, 
we call $n$-simplices $S,S' \in \calT$ \emph{strongly connected in $\calT$} if there exists a sequence $S_0=S,S_1,\dots,S'=S_m \in \calT$ such that $S_{i} \cap S_{i-1}$ is a face of both $S_{i}$ and $S_{i-1}$ for all $1 \leq i \leq m$. Clearly, strongly connected in $\calT$ is an equivalence relation. A \emph{strongly connected component} of $\calT$ is an equivalence class under this equivalence relation, and we call $\calT$ strongly connected if it has only one strongly connected component.

@@

\section{Shellable triangulations}

We return to the theory of triangulations,
as our main objective requires some further concepts.
We are interested in simplicial complexes that triangulated manifolds. 
We put particular focus on \emph{shellable} simplicial complexes as a particularly important class of complexes. Such simplicial complexes are constructed by successively adding simplices in a particularly well-behaved manner. Patches within triangulations are examples of such shellable complexes. 
Our main references for this section are the monographs by Ziegler~\cite{ziegler2012lectures} and Kozlov~\cite{kozlov2008combinatorial}. 
\\

\subsection{Triangulations of manifolds}

We define an $n$-dimensional simplicial complex to be a \emph{simplicial manifold} if the underlying set $\underlying{\calT}$ is an $n$-dimensional manifold with boundary.
We recall that this means that for every $x \in \underlying{\calT}$
there exists $\epsilon > 0$ and an embedding $\phi : \Ball_\epsilon(x) \cap \underlying{\calT} \rightarrow \bbR^n$
such that $\phi(0) = 0$ and $\phi$ is an isomorphism either onto the closed unit ball $\calB = \{ x \in \bbR^n \suchthat |x| \leq 1 \}$
or onto $\{ x \in \calB \suchthat x_1 \geq 0 \}$.
Note than any simplicial complex that triangulates an $n$-dimensional manifold must be an $n$-dimensional simplicial complex.

The following special cases receive particular interest:
a \textit{simplicial $n$-ball} is any triangulation of a topological (closed) $n$-ball,
and we sometimes call this a simplicial \textit{$n$-disk}.
A \textit{simplicial $n$-sphere} is any triangulation of a topological $n$-sphere.

We now gather a few helpful properties of simplicial manifolds,
which we state without proof.

\begin{lemma}\label{lemma:characterizationofmanifoldcomplexes}
    Let $\calT$ be a finite $n$-dimensional simplicial manifold. 
    \begin{enumerate}
        \item Any face $F \in \Faces(\calT)$ is not contained in the boundary if and only if it is contained in exactly two $n$-simplices.
        \item Any face $F \in \Faces(\calT)$ is contained in the boundary if and only if it is contained in exactly one $n$-simplex.
        % \item $\calT$ is non-branching, that is, every $F \in \calT$ of dimension $n-1$ is contained in at most two $n$-simplices of $\calT$. 
        \item If $S$ is an inner simplex, then $\patch_{\calT}(S)$ is a simplicial $n$-ball with $S$ as an inner simplex
        and $\carapace_{\calT}(S)$ is a simplicial $(n-1)$-sphere. 
        \item If $S$ is a boundary subsimplex, then $\patch_{\calT}(S)$ is a simplicial $n$-ball with $S$ as a boundary simplex,
        and $\carapace_{\calT}(S)$ is a simplicial $(n-1)$-ball.
        \item If the underlying space of $\calT$ is connected, then $\calT$ is strongly connected.
    \end{enumerate}
\end{lemma}
\begin{proof}
    % Faces have at most two parents TODO must be more careful here 
    Let $F \in \Faces(T)$ and let $z_F \in F$ be its midpoint. 
    Let $B_F$ be a ball around $z_F$ so small that it only intersects those $n$-simplices of $\calT$ that already contain $z_F$.
    Suppose there were distinct $n$-simplices $T_1, T_2, T_3$ that contain $z_F$. The intersection of any two of them is $F$, and their interiors are disjoint. 
    It follows that $B_F \setminus F$ has at least $3$ connected components, though $B_F \setminus F$ has at most two components. 
    We conclude that each face is in at most two $n$-simplices.

    % TODO inner/boundary faces

    % TODO structure of stars around simplices

    % TODO

    % TODO stars are strongly connected 
    We show that each vertex star is strongly connected via a short induction argument.
    Clearly, any simplicial $1$-ball and simplicial $1$-sphere is strongly connected. 
    Now, if $n \geq 1$, then any vertex star in an $(n+1)$-dimensional simplicial manifold is strongly connected if all simplicial $n$-balls and $n$-spheres are strongly connected.
    The induction argument implies that each vertex star in $\calT$ is strongly connected.

    % strongly connected
    If the underlying space is connected, then we easily see that the underlying space of the $1$-skeleton is path-connected.
    Given $n$-simplices $S,S' \in \calT$, 
    we can thus choose a sequence of $\hat S_0=S,\hat S_1,\dots,\hat S_m=S' \in \calT$ such that $\hat S_{i} \cap \hat S_{i-1} \neq \emptyset$ for all $1 \leq i \leq m$.
    We also notice that each vertex star is strongly connected. 
    We can thus construct change the first sequence to another $S_0=S,S_1,\dots,S'=S_m \in \calT$ 
    such that $S_{i} \cap S_{i-1}$ is a face of both $S_{i}$ and $S_{i-1}$ for all $1 \leq i \leq m$.
\end{proof}

Suppose that $\calT$ is an $n$-dimensional simplicial manifold. 
The simplices that lie in the boundary of the underlying set are called \emph{boundary simplices}.
If $S$ lies in the boundary of the underlying set, then it lies within some boundary face, % TODO
and so we conclude that the family of boundary faces constitute a simplicial complex, which we call \textit{boundary complex}.
Any point on the boundary lies within some boundary simplex, and hence the boundary complex is a triangulation of the boundary. 
Definitions imply that the boundary complex is a simplicial manifold of dimension $n-1$.

The boundary simplices of a simplicial manifold of dimension $n$ constitute a simplicial manifold themselves, of dimension $n-1$, which triangulate the boundary of the underlying space.

% We call an $n$-dimensional simplicial complex $\calT$ \emph{non-branching} if every $F \in \calT$ of dimension $n-1$ is contained in at most two $n$-simplices of $\calT$.

% TODO: https://arxiv.org/html/1607.08163v3
% https://mathoverflow.net/questions/403415/properties-a-triangulation-must-have-in-order-to-describe-a-manifold

\begin{remark}
    Clearly, not every simplicial complex is the triangulation of some (embedded) topological manifold with or without boundary. 
    Conversely, for each dimension $n \geq 5$ there exists an $n$-dimensional compact topological manifold without a triangulation. % TODO: are these compact?
    % TODO There are computational limits to algorithmically deciding whether a simplicial complex represents a manifold. 
\end{remark}
\begin{remark}
    Whether a simplicial complex is the triangulation of a manifold is algorithmically undecidable. 
    For that reason, there is no easy combinatorial property that indicates whether a simplicial complex triangulates a manifold.
\end{remark}

% TODO: the following lemma needs only be as strong as it aids the subsequent discussion

@@

    \caption{Example of a connected $2$-dimensional simplicial manifold, which triangulates an annulus, that does neither admit a manifold-like $0$-patching nor manifold-like $1$-patching.}\label{figure:annuluscounterexample}
\end{figure}

@@

We turn our attention to a different idea of how to incrementally construct a simplicial complex while maintaining well-behaved intermediate complexes. 
We build upon the notion of shelling, as introduced in~\cite[Definition 8.1]{ziegler2012lectures}.
Our definition of shelling is equivalent to the notion of the shellings of simplicial complexes in the aforementioned reference; see also~\cite[Remark~8.3]{ziegler2012lectures}.

Suppose that $\calT$ is an $n$-dimensional simplicial complex and we have an enumeration of the $n$-simplices $T_{0}, T_{1}, T_{2}, \dots \in \subsimplex_{n}(\calT)$.
For any enumeration, we call 
\[
    I_m := ( \patch_{\calT}(T_{0}) \cup \patch_{\calT}(T_{1}) \cup \dots \cup \patch_{\calT}(T_{m}) ) \cap \patch_{\calT}(T_{m+1})
\]
be the $m$-th \textit{interface complex}. Note that this intersection is automatically a subcomplex of the boundary complex of the last simplex $T_{m}$.
We call the enumeration a \emph{shelling} if each interface $I_m$ is a simplicial complex of dimension $n-1$.

\begin{remark}
    We interpret a shelling as the construction of a triangulation 
    by successively attaching simplices such that the intermediate triangulations are well-behaved. 
    Conversely, the reverse enumeration describes a successive decomposition of the triangulation, hence the name ``shelling''.
\end{remark}

\begin{example}
    Any simplex $T$ trivially has a shelling. The boundary complex $\partial \calT(T)$ has a shelling:
    in fact, any enumeration of the boundary faces of $T$ constitutes a shelling.
\end{example}
\begin{example}
    The standard triangulation of the $n$-dimensional cube, the Kuhn triangulation, is shellable. 
\end{example}
\begin{example}
    Not all triangulable sets admit a triangulation that is also shellable. 
    Moreover, even if a set admits a shellable triangulation, not all of its triangulations might be shellable. 
    We refer to~\cite[Example 8.9]{ziegler2012lectures} for an examples of non-shellable triangulations of cubes and tetrahedra in three dimensions. 
\end{example}

Our main interest are the local patches or stars within triangulations. 
It turns out that those are shellable.

\begin{lemma}
    Any simplicial $2$-disk is shellable.
    Any simplicial $2$-sphere is shellable. 
\end{lemma}
\begin{proof}
    Let $\calT$ be a simplicial $2$-disk. 
    If $\calT$ contains only one triangle, then there is nothing to show. 
    Otherwise, there exists at least one triangle $T$ that touches the boundary of the disk at either one or two edges {\color{red}Needs proof}. 
    Letting $\calT'$ be the triangulation obtained by removing $T$,
    we have another triangulation of the $2$-disk that intersects $T$ only at either two or one edges.
    Repeating this procedure yields an enumeration of the triangles of $\calT$,
    whose reverse is a shelling.

    Now let $\calS$ be any simplicial $2$-sphere. 
    By removing any triangle $S \in \calS$, we obtain a simplicial $2$-disk, which is shellable. 
    Any of its shellings can be extended to a shelling of $\calS$ by re-inserting the first triangle $S$.
\end{proof}

\begin{lemma}
    Let $\calT$ be a $n$-dimensional simplicial manifold and $V \in \Vertices(\calT)$ be an inner vertex.
    Then $\patch_{\calT}(V)$ is shellable. 
\end{lemma}
\begin{proof}
    Without loss of generality, $V$ is the origin. 
    Any ray starting at $V$ intersects at most one vertex of $\patch_{\calT}(V)$.
    We can also assume that all boundary vertices of $\patch_{\calT}(V)$ to be unit vectors,
    which does not affect shellability. 
    The simplicial boundary complex of $\calT$ is the boundary complex of a convex polytope,
    and is thus shellable~\cite{ziegler2012lectures}. % TODO: cite exactly 
\end{proof}

\begin{lemma}
    Let $\calT$ be a $3$-dimensional simplicial manifold and $S \in \calT$.
    Then $\patch_{\calT}(S)$ is shellable. 
\end{lemma}
\begin{proof}
    The statement is clear if $S$ is an inner or boundary face of $\calT$. 
    The statement is also easily verified if $S$ is an inner or boundary edge of $\calT$.
    When $S$ is an inner vertex, then we apply the preceding lemma.

    It remains to consider the case where $S$ is a boundary vertex. 
    Then the triangles of $\patch_{\calT}(S)$ that do not contain $V$ constitute a simplicial $2$-disk {\color{red}Needs proof},
    which is shellable,
    and any such shelling can be extended to a shelling of $\patch_{\calT}(S)$.
\end{proof}

\begin{remark}
    Patches around boundary simplices are not necessarily shellable when the dimension exceeds $3$.
    For example, there exists a non-shellable triangulation of a tetrahedron. 
    If we extend this triangulation to a triangulation of a hypertetrahedron by suspending it from a new point $v_\star$, 
    then the resulting new triangulation is non-shellable and coincides with the patch around $v_\star$.
\end{remark}

The reason of our interest in shellable simplicial complexes is that they can be constructed by successive adhesion of simplices to intermediate complexes, 
such that the interface is well-behaved: it is a simplicial disk or a simplicial sphere. 
That has the following important consequence.

% TODO: we only need some basic properties, namely that the construction maintains manifolds with boundary.
% TODO: copy and adapt the content of Ziegler and Kozlov as necessary

\begin{lemma}
    If an $n$-dimensional simplicial complex $\calT$ has a shelling $T_{0}, T_{1}, T_{2}, \dots$, then it is a simplicial manifold.
    In particular, $\patch_{\calT}(T_{0}) \cup \patch_{\calT}(T_{1}) \cup \dots \cup \patch_{\calT}(T_{m})$ is a simplicial manifold for every $m \geq 0$.
\end{lemma}
\begin{proof}    
\end{proof}

@@

\begin{remark}
    Whether a simplicial complex is shellable can be checked, in principle, simply by trying out all the possibly enumerations.
    That we cannot do much better than this is captured in the result that testing for shellability is NP-complete~\cite{goaoc2019shellability}:
    this complexity result is even true if we merely consider simplicial complexes of dimension two embedded in some Euclidean space.
\end{remark}

\subsection{On the topology of shellable triangulations}

A shellable simplicial complex that triangulates a manifold has a very restricted topology. 
The Mayer-Vietoris theorem implies restrictions on the homology groups of simplicial complexes iteratively constructed via shelling.

@@

% then $H_{n}(\calT') \simeq H_{n}(\calT \oplus T)$ and $H_{n-1}(\calT') \simeq H_{n-1}(\calT \oplus T)$ follow immediately.
% Moreover, Thus $H_{n}(\calT') = H_{n}(\calT) \oplus H_{n}(\calS)$.
% We conclude that any shellable simplicial complex has the homology groups of a disk or a sphere. 
% NB: the zeroth homology is always free, and the n-th homology group, being the kernel of a mapping free abelian group, is free as well.
% \begin{align*}
%     \begin{CD}
%         0 \longrightarrow H_{1}(\calT) \oplus H_{1}(T) \longrightarrow H_{1}(\calT') \longrightarrow H_{0}(\calS) \longrightarrow H_{0}(\calT) \oplus H_{0}(T) \longrightarrow H_{0}(\calT') \longrightarrow 0 
%     \end{CD}
% \end{align*}
